%% datacube-rs — Computers & Geosciences software paper
%% Elsevier elsarticle template. Compile: latexmk -pdf datacube-rs.tex
%%   (or: pdflatex; bibtex datacube-rs; pdflatex; pdflatex)
\documentclass[review]{elsarticle}

\usepackage{lineno,hyperref}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{tikz}
\usetikzlibrary{arrows.meta,positioning,fit,backgrounds}
\modulolinenumbers[5]

%% Fig. 1 styles
\tikzset{
  stage/.style={draw, rounded corners=2pt, align=center, font=\small,
    inner sep=4pt, minimum height=8mm, fill=blue!4},
  core/.style={stage, fill=green!6},
  outbox/.style={stage, fill=orange!6},
  flow/.style={-{Stealth[length=2mm]}, thick},
}

\journal{Computers \& Geosciences}

%% Numbered reference style with author names available for textual cites (C&G)
\bibliographystyle{elsarticle-num-names}

\begin{document}

\begin{frontmatter}

\title{datacube-rs: An open-source Rust engine for streaming temporal
analysis of remote-sensing data cubes}

\author[usach]{Francisco Parra}
\ead{francisco.parra.o@usach.cl}
\address[usach]{Universidad de Santiago de Chile (USACH), Santiago, Chile}

\begin{abstract}
Open satellite archives have made remote sensing a temporal discipline, but the
tools that assemble image collections into space--time data cubes ship only
generic reducers, leaving trend, phenology and structural-break analysis to
separate single-purpose packages in different languages. We present
\texttt{datacube-rs}, an open-source engine written in Rust that unifies
temporal data-cube analysis with a validated set of per-pixel temporal
estimators in a single, portable artifact. It models a cube on the
$(\text{band}, y, x, \text{time})$ axes with the time dimension innermost,
enabling zero-copy per-pixel and per-chunk streaming parallelised with Rayon;
ingests satellite time series directly from STAC catalogues and Cloud-Optimized
GeoTIFF assets with cross-UTM-zone mosaicking; and provides ordinary-least-squares
trend, the Theil--Sen slope and tie-corrected Mann--Kendall test, harmonic
regression for seasonality, and OLS-CUSUM structural-break detection. Every
estimator is cross-validated to numerical parity against established reference
implementations (pyMannKendall, SciPy and statsmodels), passing 103 of 103
checks at a relative tolerance of $10^{-9}$. The same compiled core is delivered
as a native Rust library, a command-line tool, a Python extension (PyO3) and a
WebAssembly module that runs in the browser. We report parallel-scaling
benchmarks and a multiannual Sentinel-2 case study.
\end{abstract}

\begin{keyword}
Data cube \sep Satellite image time series \sep Rust \sep STAC \sep
Mann--Kendall \sep Phenology \sep Structural breaks
\end{keyword}

\end{frontmatter}

\linenumbers

\section{Introduction}
\label{sec:intro}

Open satellite archives have turned remote sensing into a fundamentally temporal
discipline. The full Landsat record and the Sentinel missions are now
distributed as free, analysis-ready data (ARD) through cloud object storage, and
a single study area may be covered by hundreds of observations per year. The
scientific questions that follow are, correspondingly, questions about time: is a
pixel's vegetation index trending up or down, and is that trend significant; when
does the growing season start and how is its amplitude changing; and where and
when does an abrupt disturbance---a fire, a clear-cut, a landslide---break the
expected seasonal signal. Answering them at scale requires both an efficient way
to assemble irregularly sampled, cloud-masked image collections into a regular
space--time cube, and a set of per-pixel statistical estimators applied along the
temporal axis.

A mature ecosystem addresses the first need. The \emph{gdalcubes} library
\citep{appel2019gdalcubes} builds on-demand regular data cubes from image
collections through GDAL, and endorses Cloud-Optimized GeoTIFF (COG) for
efficient partial reads. The Open Data Cube \citep{lewis2017agdc} provides a
Python framework for indexing and managing petabyte-scale ARD archives,
demonstrated on national high-performance infrastructure. The \emph{openEO} API
\citep{schramm2021openeo} harmonises heterogeneous cloud back-ends behind a
single, STAC-aligned virtual data-cube interface. \emph{xarray}
\citep{hoyer2017xarray} supplies the labeled $N$-dimensional array model that
underpins much of the scientific Python stack, and Google Earth Engine
\citep{gorelick2017gee} offers planetary-scale analysis to anyone with a browser.
Together these tools have made cube construction, ARD management and cloud access
broadly available.

The second need---temporal statistical analysis---is comparatively fragmented.
Cube engines expose generic reducers and aggregations (mean, median, percentile,
moving windows) but leave domain estimators to the user: gdalcubes, the Open Data
Cube, openEO and xarray ship no built-in monotonic-trend test, robust slope,
harmonic phenology model or structural-break detector, deferring these to
user-defined functions or external libraries. Google Earth Engine is a partial
exception---it provides Sen's slope and linear/robust regression reducers---but
it is a proprietary service whose computation cannot run offline and into whose
API existing workflows must be rewritten. Among cube tools, FORCE
\citep{frantz2019force} goes furthest, adding land-surface phenology and linear
trend analysis to a local ARD store, but it is a heavy C/C++ and
high-performance-computing stack that writes to a bespoke local grid rather than
streaming from cloud-native catalogues. Meanwhile the reference temporal methods
themselves live in separate, single-purpose packages and languages: the
Mann--Kendall trend-test family and Theil--Sen slope in \emph{pyMannKendall}
\citep{hussain2019pymannkendall}; the harmonic-plus-trend seasonality model in
Continuous Change Detection and Classification (CCDC; \citealp{zhu2014ccdc}); and
breakpoint detection on decomposed series in BFAST \citep{verbesselt2010bfast}. A
researcher who wants a significance-tested trend map, a phenology fit and a
disturbance-timing map from the same cube must therefore cross several tools,
languages and runtimes, and must trust that each per-pixel estimator is correctly
implemented.

Here we present \texttt{datacube-rs}, an open-source engine, written in Rust,
that unifies temporal data-cube analysis and a validated set of per-pixel
temporal estimators in a single, portable artifact. \texttt{datacube-rs} models a
cube on the $(\text{band}, y, x, \text{time})$ axes with the time dimension
innermost, so that each pixel's series is a contiguous slice and can be
streamed---per pixel or by spatial chunk---without copying; per-pixel computation
is parallelised across cores with Rayon. It ingests scenes directly from STAC
catalogues and COG assets (reusing the cloud stack of the SurtGIS engine),
mosaicking across UTM zones onto a common grid. On top of the cube it provides
ordinary-least-squares trend, the Theil--Sen robust slope and the tie-corrected
Mann--Kendall test, harmonic (Fourier) regression for seasonality and phenology,
and structural-break detection by an OLS-CUSUM test with recursive binary
segmentation---the latter inspired by BFAST's decompose-then-detect strategy,
though it deliberately uses a different test (OLS-CUSUM rather than BFAST's
OLS-MOSUM with Bai--Perron selection; see Section~\ref{sec:methods}). Every
estimator treats missing values as \texttt{NaN} and drops them pairwise, and
every reported quantity is cross-validated to numerical parity against
established reference implementations---pyMannKendall, SciPy and
statsmodels---passing 103 of 103 checks at a relative tolerance of
$10^{-9}$. The same compiled core is exposed as a native Rust library, a
command-line tool, a Python extension (via PyO3) and a WebAssembly module that
runs the estimators in the browser.

The contributions of this paper are:
\begin{itemize}
  \item a streaming $(\text{band}, y, x, \text{time})$ cube model whose
  contiguous time axis enables zero-copy per-pixel and per-chunk iteration,
  parallelised with Rayon;
  \item a set of per-pixel temporal estimators---OLS trend, Theil--Sen,
  tie-corrected Mann--Kendall, harmonic regression and OLS-CUSUM structural
  breaks---implemented natively and validated to $10^{-9}$ numerical parity
  against pyMannKendall, SciPy and statsmodels;
  \item direct ingestion of satellite time series from STAC/COG with
  cross-UTM-zone mosaicking, plus temporal compositing and gap-filling on the
  cube;
  \item a single dependency-light Rust core delivered as four deployment
  targets---library, CLI, Python and in-browser WebAssembly---with reproducible
  benchmarks of its parallel scaling.
\end{itemize}

The remainder of the paper is organised as follows.
Section~\ref{sec:methods} states the temporal estimators and their reference
implementations. Section~\ref{sec:engine} describes the \texttt{datacube-rs}
engine: the cube data model, the streaming iterator and chunking, STAC/COG
ingestion and the deployment targets. Section~\ref{sec:validation} reports the
numerical-parity validation, performance benchmarks and a real Sentinel-2 case
study. Section~\ref{sec:discussion} discusses how \texttt{datacube-rs} relates to
existing tools and its current limitations, and Section~\ref{sec:conclusions}
concludes.

\section{Temporal estimators}
\label{sec:methods}

\texttt{datacube-rs} operates on the time series of a single pixel and band,
$\{(t_i, y_i)\}_{i=1}^{n}$, where $t_i$ is a numeric time coordinate (e.g.\
fractional years) and $y_i$ the observed value. Cloud-masked or otherwise
missing observations are encoded as \texttt{NaN}; every estimator first drops
the non-finite pairs and operates on the remaining finite observations, so that
the irregular sampling produced by cloud screening is handled without
interpolation. Because $t_i$ enters the estimators explicitly, unevenly spaced
series are treated correctly. The estimators reproduce established methods; the
contribution of this paper is their unified, validated native implementation
(Section~\ref{sec:engine}--\ref{sec:validation}), not the methods themselves.

\subsection{Linear trend (ordinary least squares)}
\label{sec:ols}
The least-squares line $y = a + b\,t$ has slope and intercept
\begin{equation}
b = \frac{\sum_i (t_i - \bar t)(y_i - \bar y)}{\sum_i (t_i - \bar t)^2},
\qquad
a = \bar y - b\,\bar t,
\end{equation}
with $\bar t, \bar y$ the means. We report the coefficient of determination
$R^2 = 1 - \mathrm{SS_{res}}/\mathrm{SS_{tot}}$, the standard error of the slope
$\mathrm{se}(b) = \sqrt{\mathrm{SS_{res}}/(n-2)\big/\sum_i (t_i-\bar t)^2}$, and a
two-sided $p$-value for $H_0\!: b = 0$ from Student's $t$ statistic
$b/\mathrm{se}(b)$ with $n-2$ degrees of freedom. The OLS slope is efficient but
sensitive to outliers, which motivates the robust estimators below.

\subsection{Theil--Sen robust slope}
\label{sec:theilsen}
The Theil--Sen estimator \citep{theil1950,sen1968} is the median of the slopes
of all observation pairs,
\begin{equation}
b_{\mathrm{TS}} = \operatorname*{median}_{\,t_i \neq t_j}
  \frac{y_j - y_i}{t_j - t_i},
\qquad
a_{\mathrm{TS}} = \operatorname{median}(y) - b_{\mathrm{TS}}\,\operatorname{median}(t),
\end{equation}
evaluated over the $\binom{n}{2}$ pairs with distinct time coordinates. It has a
breakdown point of about $29\%$ and is the slope estimator conventionally paired
with the Mann--Kendall test.

\subsection{Mann--Kendall trend test}
\label{sec:mk}
The Mann--Kendall test \citep{mann1945,kendall1975} assesses the presence of a
monotonic trend through the statistic
\begin{equation}
S = \sum_{i<j} \operatorname{sgn}(y_j - y_i),
\end{equation}
where $\operatorname{sgn}$ returns $-1,0,+1$ (ties contribute $0$). For $n$
observations containing tie groups of sizes $t_p$, the tie-corrected variance is
\begin{equation}
\label{eq:varS}
\operatorname{Var}(S) = \frac{1}{18}\!\left[ n(n-1)(2n+5)
  - \sum_{p} t_p (t_p - 1)(2 t_p + 5) \right],
\end{equation}
and the continuity-corrected standard normal score is
\begin{equation}
Z =
\begin{cases}
(S-1)/\sqrt{\operatorname{Var}(S)}, & S > 0,\\[2pt]
0, & S = 0,\\[2pt]
(S+1)/\sqrt{\operatorname{Var}(S)}, & S < 0.
\end{cases}
\end{equation}
We report Kendall's $\tau = S/\binom{n}{2}$ and the two-sided $p$-value
$p = 2\,[1-\Phi(|Z|)]$, with $\Phi$ the standard normal CDF; at significance
level $\alpha$ the trend is declared increasing ($Z>0$) or decreasing ($Z<0$)
when $p<\alpha$, and absent otherwise. This is the ``original'' tie-corrected
test as implemented in \texttt{pyMannKendall} \citep{hussain2019pymannkendall},
against which our implementation is validated to numerical parity
(Section~\ref{sec:validation}); we note that the variance formula
\eqref{eq:varS} originates with \citet{mann1945,kendall1975}, the JOSS note being
the parity target rather than the source of the equations.

\subsection{Harmonic regression}
\label{sec:harmonic}
Seasonality and phenology are described by a Fourier-series model with a linear
trend term,
\begin{equation}
\label{eq:harmonic}
\hat y(t) = a + b\,t
  + \sum_{k=1}^{K} \Big[ \alpha_k \cos\!\Big(\tfrac{2\pi k t}{P}\Big)
                       + \beta_k  \sin\!\Big(\tfrac{2\pi k t}{P}\Big) \Big],
\end{equation}
where $P$ is the season length (e.g.\ one year) and $K$ the number of harmonics.
The coefficients are obtained by ordinary least squares, solving the normal
equations $X^{\!\top}\!X\,\boldsymbol\beta = X^{\!\top} y$; the linear term is
centred on $\bar t$ for numerical conditioning, and the design matrix is
checked for rank deficiency (e.g.\ a harmonic aliased by the sampling). Each
harmonic is summarised by its amplitude and phase,
\begin{equation}
A_k = \sqrt{\alpha_k^2 + \beta_k^2}, \qquad
\varphi_k = \operatorname{atan2}(\beta_k, \alpha_k),
\end{equation}
so that the $k$-th component peaks at $t = P\varphi_k/(2\pi k)$ within its
period---directly interpretable as phenological timing and magnitude. The single
annual harmonic ($K=1$) with a linear inter-annual term is the seasonality model
of CCDC \citep{zhu2014ccdc}.

\subsection{Structural break detection}
\label{sec:breaks}
Abrupt changes are detected in the spirit of BFAST
\citep{verbesselt2010bfast}---fit a trend-plus-season model, then test its
residuals for instability---but with a different test. For a candidate segment we
fit the model \eqref{eq:harmonic} (optionally trend-only, $K=0$) by OLS, obtain
residuals $e_i$, and form the OLS-CUSUM process \citep{brown1975,ploberger1992}
\begin{equation}
W_k = \frac{1}{\hat\sigma \sqrt{n}} \sum_{i=1}^{k} e_i,
\qquad
\hat\sigma^2 = \frac{1}{n-p}\sum_{i=1}^{n} e_i^2,
\qquad k = 1,\dots,n,
\end{equation}
with $p = 2 + 2K$ the number of fitted parameters. The test statistic is the
supremum $M = \max_k |W_k|$. Under the null of parameter constancy, $W$ converges
to a standard Brownian bridge, whose supremum has the Kolmogorov distribution,
giving the $p$-value
\begin{equation}
\label{eq:bb}
\Pr\!\big(\sup|W| > M\big)
  = 2 \sum_{j=1}^{\infty} (-1)^{j+1} \exp(-2 j^2 M^2),
\end{equation}
identical to \texttt{scipy.stats.kstwobign} and to
\texttt{statsmodels} \texttt{breaks\_cusumolsresid}, against which it is validated
(Section~\ref{sec:validation}). When a segment tests significant at level
$\alpha$, a break is placed at $\arg\max_k |W_k|$, constrained so that both sides
retain at least a minimum number of observations, and the procedure recurses on
each sub-segment (binary segmentation), yielding the set of break times.

This differs deliberately from BFAST in two respects, which we state explicitly
because readers familiar with BFAST will expect them. First, BFAST tests for
instability with an OLS-MOSUM (moving-sum) statistic rather than the OLS-CUSUM
used here. Second, BFAST estimates the number and position of breaks by the
dynamic-programming procedure of \citet{baiperron2003} with the number of breaks
chosen by an information criterion, whereas \texttt{datacube-rs} uses greedy
binary segmentation. Our choice yields a single, closed-form $p$-value per
segment \eqref{eq:bb} and an $O(n)$ test that suits per-pixel evaluation over
large cubes; it is validated against the OLS-CUSUM reference implementation, not
against the R \texttt{bfast} package.

\section{The datacube-rs engine}
\label{sec:engine}

\texttt{datacube-rs} is a Rust workspace (edition 2024) organised as a compute
core with no input/output, a cloud-ingestion crate, and three thin frontends.
The core, \texttt{datacube-core}, holds the cube data model, the streaming
iterators, and the estimators of Section~\ref{sec:methods}; it depends only on
\texttt{ndarray}, \texttt{rayon}, and \texttt{libm}, and implements the
least-squares normal equations and the special functions (error and incomplete
beta functions for the $p$-values) directly, with no linear-algebra dependency.
Figure~\ref{fig:arch} shows the dataflow from satellite scenes to per-pixel
results and the frontends that wrap the shared core.

\begin{figure}[t]
\centering
\begin{tikzpicture}[node distance=4.5mm]
  \node[stage] (src) {STAC catalogue $+$ COG assets\\\footnotesize (Planetary Computer / Earth Search)};
  \node[stage, below=of src] (io) {\texttt{datacube-io}: search $\rightarrow$ cloud filter\\$\rightarrow$ windowed COG read $\rightarrow$ UTM reproject\\$+$ cross-zone mosaic $\rightarrow$ grid align};
  \node[core, below=of io] (cube) {\textbf{Cube} $(\mathrm{band}, y, x, t)$\\\footnotesize contiguous time axis, \texttt{NaN} nodata};
  \node[core, below=of cube] (iter) {Streaming iterators\\\footnotesize \texttt{iter\_series} $\cdot$ \texttt{par\_map\_series} (Rayon) $\cdot$ \texttt{chunks}};
  \node[core, below=of iter, text width=82mm] (ops) {Estimators: OLS $\cdot$ Theil--Sen $\cdot$ Mann--Kendall $\cdot$ harmonic $\cdot$ OLS-CUSUM\\[2pt]Transforms: temporal composite $\cdot$ linear gap-fill};
  \node[outbox, below=of ops] (outp) {Outputs: GeoTIFF trend / break maps $\cdot$ JSON reports};

  \draw[flow] (src) -- (io);
  \draw[flow] (io) -- (cube);
  \draw[flow] (cube) -- (iter);
  \draw[flow] (iter) -- (ops);
  \draw[flow] (ops) -- (outp);

  \begin{scope}[on background layer]
    \node[draw, dashed, rounded corners, fit=(cube)(iter)(ops),
      inner sep=6pt, label={[font=\footnotesize\itshape]right:datacube-core}] {};
  \end{scope}

  \node[stage, right=14mm of iter, text width=24mm, fill=gray!6] (tgt)
    {\footnotesize Frontends:\\Rust lib\\CLI\\Python (PyO3)\\WASM};
  \draw[flow, dashed] (tgt) -- (iter);
\end{tikzpicture}
\caption{Architecture of \texttt{datacube-rs}. Scenes are ingested from STAC/COG
by \texttt{datacube-io}, assembled into a $(\mathrm{band}, y, x, t)$ cube whose
innermost time axis makes each pixel series contiguous, streamed through the
per-pixel/chunk iterators, and reduced by the temporal estimators and
transforms. The compute core (dashed) is exposed unchanged through four
frontends.}
\label{fig:arch}
\end{figure}

\subsection{Cube data model}
\label{sec:model}
A cube is a four-dimensional array on the axes $(\mathrm{band}, y, x, t)$ with
the time dimension \emph{innermost}, paired with a numeric time coordinate
vector and a list of band names. Placing time last means that the complete
series of a pixel occupies a contiguous region of memory: it can be borrowed as
a slice and passed to an estimator with no copy and no gather. The constructor
validates that the coordinate lengths match the array extents and, if necessary,
copies the data into standard (row-major) layout so the contiguity guarantee
holds. Missing observations are represented as \texttt{NaN}, which the estimators
drop pairwise (Section~\ref{sec:methods}); no separate mask array is carried.

\subsection{Streaming iteration and parallel evaluation}
\label{sec:streaming}
Three access patterns are built on the contiguous layout. \texttt{iter\_series}
yields every pixel series of a band in row-major order as a zero-copy view;
\texttt{chunks} yields spatial tiles (all bands and times) for blocked
processing; and \texttt{par\_map\_series}---the central primitive for per-pixel
analysis---applies a closure $f(\mathbf{t}, \mathbf{y})$ to every pixel series in
parallel with Rayon and collects the results into a $(y, x)$ grid. A trend map is
thus expressed as a single call whose closure invokes the relevant estimator;
because the worker threads read disjoint contiguous slices and write disjoint
output cells, no locking is required. Section~\ref{sec:validation} shows that
throughput scales with the core count, so the cost is set by the per-pixel
estimator rather than by parallel overhead.

\subsection{Temporal compositing and gap-filling}
\label{sec:transforms}
Two cube-to-cube transforms support pre-processing. Compositing groups time
slices---either those sharing a time coordinate (e.g.\ adjacent satellite tiles
acquired in the same overpass) or fixed-width temporal bins such as months---and
reduces each group per pixel with a \texttt{NaN}-aware median, mean, minimum, or
maximum; the median composite is the standard cloud-robust reduction. Linear
gap-filling interpolates \texttt{NaN} runs along time from the nearest finite
observations on each side, leaving gaps wider than a configurable limit (and the
series ends) untouched, so that no value is extrapolated. Both transforms run in
parallel over pixels.

\subsection{Cube ingestion from STAC and COG}
\label{sec:ingest}
The \texttt{datacube-io} crate assembles a cube directly from a satellite
archive. Given a collection, a WGS84 bounding box, a date range, and a list of
asset (band) keys, it queries a STAC API---Microsoft Planetary Computer, Element
84 Earth Search, or any STAC endpoint---optionally discarding scenes above a
cloud-cover threshold. For each retained scene it signs the asset URL when
required (Planetary Computer shared-access signatures), reads the Cloud-Optimized
GeoTIFF windowed to the bounding box (optionally at a coarser overview level for
speed), and masks the nodata value to \texttt{NaN}. The first scene read defines
a reference grid; subsequent scenes are reprojected and resampled onto it, and
scenes lying in a neighbouring UTM zone are mosaicked across the zone boundary
rather than discarded. Acquisition timestamps are converted to fractional years
so that an annual cycle has unit period, matching the harmonic model of
Section~\ref{sec:harmonic}. The heavy machinery---HTTP range reads, TIFF
decoding, signature handling, and UTM reprojection---is reused from the SurtGIS
engine's cloud stack, keeping \texttt{datacube-io} a thin assembly layer; it is
an optional build feature so that the core and frontends compile without it.

\subsection{Deployment targets}
\label{sec:targets}
The same compiled core is delivered through four frontends. As a \emph{Rust
library} it is embedded directly in other tools. The \emph{command-line}
program exposes the per-series estimators on CSV input and, with the ingestion
feature enabled, a \texttt{stack} command that searches, assembles, and writes
GeoTIFF trend- and break-map products in one step. The \emph{Python} extension
(built with PyO3) presents the estimators and a cube class over NumPy arrays;
because the Rayon workers touch no Python objects, the global interpreter lock is
released during per-pixel evaluation, so a trend map of a moderate cube completes
in well under a second from Python (Section~\ref{sec:validation}). The
\emph{WebAssembly} module exposes the per-series estimators to JavaScript and
powers an in-browser demonstration that fits a harmonic model and detects breaks
on a synthetic series with no server-side computation. All targets share one
implementation, so a result obtained at the command line, in a Python notebook,
or in a browser is computed by identical code.

\section{Validation and benchmarking}
\label{sec:validation}

We validate \texttt{datacube-rs} in three ways: numerical agreement with
established reference implementations of each estimator, performance benchmarks
of the per-pixel estimators and the parallel iterator, and an end-to-end case
study on a real Sentinel-2 cube.

\subsection{Numerical parity}
\label{sec:parity}
Because the estimators reproduce known methods, correctness is established by
agreement with their reference implementations rather than against a ground
truth. A validation harness generates deterministic test series---monotonic,
tied, constant, noisy, seasonal, and gap-containing---and compares every
reported quantity of \texttt{datacube-rs} against the corresponding value from a
reference. Trend and slope are checked against \texttt{pyMannKendall}
\citep{hussain2019pymannkendall} and \texttt{scipy.stats.linregress}; the
harmonic fit against \texttt{numpy.linalg.lstsq} on the same design matrix; and
the OLS-CUSUM break statistic and $p$-value against
\texttt{statsmodels} \texttt{breaks\_cusumolsresid}. Table~\ref{tab:parity}
summarises the result: all 103 checks agree to within a relative tolerance of
$10^{-9}$.

\begin{table}[t]
\centering
\caption{Numerical-parity validation. Each estimator is compared to its
reference implementation over deterministic test series; every reported field
(coefficients, statistics and $p$-values) is one check. All agree to a relative
tolerance of $10^{-9}$.}
\label{tab:parity}
\small
\begin{tabular}{@{}llcc@{}}
\toprule
Estimator & Reference & Series & Checks \\
\midrule
Mann--Kendall      & \texttt{pyMannKendall.original\_test} & 7 & 42 \\
Theil--Sen         & \texttt{pyMannKendall.sens\_slope}    & 7 & 13 \\
OLS trend          & \texttt{scipy.stats.linregress}       & 6 & 30 \\
Harmonic regression& \texttt{numpy.linalg.lstsq}           & 1 & 12 \\
OLS-CUSUM breaks   & \texttt{statsmodels.breaks\_cusumolsresid} & 3 & 6 \\
\midrule
\textbf{Total}     &                                       &   & \textbf{103} \\
\bottomrule
\end{tabular}
\end{table}

Two intentional divergences from the references are worth noting, as they would
otherwise read as discrepancies. For a constant series \texttt{scipy} returns
\texttt{NaN} for the coefficient of determination, standard error and $p$-value
of the regression, where \texttt{datacube-rs} defines the (degenerate) perfect
fit $R^2 = 1$, $p = 1$; and \texttt{pyMannKendall}'s Sen-slope assumes unit time
spacing after dropping missing values, whereas \texttt{datacube-rs} retains the
true time gaps, so for the gap-containing series the Theil--Sen slope is compared
against the median of pairwise slopes computed on the real coordinates.

\subsection{Performance}
\label{sec:perf}
Benchmarks were run with Criterion on a 12th-generation Intel Core i7-1270P
(16~logical cores) under Rust~1.94. Absolute timings are machine-dependent; the
relevant quantities are the scaling with series length and with pixel count.
Table~\ref{tab:perseries} reports the per-series cost of each estimator for
series of $50$, $230$ and $1000$ observations, spanning roughly two years of
Sentinel-2 revisit to a dense daily record. The linear and harmonic fits are
$O(n)$ and remain in the microsecond range; Theil--Sen and Mann--Kendall are
$O(n^2)$ in the number of finite observations, as expected from the pairwise
construction, and dominate for long records.

\begin{table}[t]
\centering
\caption{Per-series estimator cost (median) for $n = 50$, $230$ and $1000$
observations.}
\label{tab:perseries}
\small
\begin{tabular}{@{}lccc@{}}
\toprule
Estimator & $n{=}50$ & $n{=}230$ & $n{=}1000$ \\
\midrule
OLS trend            & $1.4\,\mu$s & $3.8\,\mu$s & $13\,\mu$s \\
Mann--Kendall        & $3.4\,\mu$s & $51\,\mu$s  & $1.0\,$ms \\
Theil--Sen           & $26\,\mu$s  & $120\,\mu$s & $5.2\,$ms \\
Harmonic ($K{=}2$)   & $13\,\mu$s  & $188\,\mu$s & $440\,\mu$s \\
OLS-CUSUM breaks     & $16\,\mu$s  & $234\,\mu$s & $412\,\mu$s \\
\bottomrule
\end{tabular}
\end{table}

At the cube level, the \texttt{par\_map\_series} primitive was benchmarked on
cubes of $64^2$, $128^2$ and $256^2$ pixels over $60$ time steps
(Table~\ref{tab:scaling}). Wall-clock time grows in proportion to the pixel
count---throughput stays within the same band as the grid grows sixteen-fold,
from $4$k to $65$k pixels---confirming that the iterator scales linearly with
the Rayon pool and that the cost is set by the per-pixel estimator rather than by
parallel overhead. A Theil--Sen-plus-Mann--Kendall trend map of a $256^2$ cube
completes in about $0.6$\,s on this laptop. The temporal transforms over a
$128^2\times120$ cube take $11$\,ms (median monthly composite) and $12$\,ms
(linear gap-fill); compositing parallelises well, whereas gap-filling is
memory-bandwidth bound and gains little from threading.

\begin{table}[t]
\centering
\caption{Wall-clock time of \texttt{par\_map\_series} over $(1, n_y, n_x, 60)$
cubes, Rayon-parallel across 16 cores.}
\label{tab:scaling}
\small
\begin{tabular}{@{}lccc@{}}
\toprule
Estimator & $64^2$ & $128^2$ & $256^2$ \\
\midrule
Theil--Sen     & $44\,$ms & $116\,$ms & $581\,$ms \\
Mann--Kendall  & $33\,$ms & $70\,$ms  & $326\,$ms \\
Harmonic ($K{=}2$) & $51\,$ms & $140\,$ms & $808\,$ms \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Case study: a cross-zone Sentinel-2 cube}
\label{sec:casestudy}
To exercise the full ingestion-to-analysis path we built a cube over a small
area on the boundary between UTM zones 18S and 19S near Santiago, Chile
($-72^\circ$ longitude), from the Sentinel-2 L2A collection on the Microsoft
Planetary Computer. Querying the red band over the first half of 2024 with a
$50\%$ cloud-cover threshold returns scenes tiled in both zones. With cross-zone
mosaicking enabled, \texttt{datacube-rs} reprojects the neighbouring-zone tiles
onto the reference grid and assembles $31$ usable time slices ($15$ from zone~19
and $16$ reprojected from zone~18); with mosaicking disabled the same query
yields only $15$ slices, the zone-18 tiles being skipped. Cross-zone mosaicking
thus doubles the usable temporal depth for an area that straddles a zone
boundary---a common situation for study areas that are not chosen to fit the
satellite tiling. From the resulting cube, per-pixel Theil--Sen slope,
Mann--Kendall significance and OLS-CUSUM break maps are produced and written as
georeferenced GeoTIFFs in a single command. This is a demonstration of the
workflow rather than a domain study; the estimators it applies are those
validated in Section~\ref{sec:parity}.

\section{Discussion}
\label{sec:discussion}
% TODO: positioning vs gdalcubes/ODC/openEO/GEE/FORCE; what Rust single-binary
% + WASM buys; limitations + roadmap (GeoZarr backing, compositing).

\section{Conclusions}
\label{sec:conclusions}
% TODO

%% --- Back-matter (Elsevier / C&G order) ---
\section*{CRediT authorship contribution statement}
\textbf{Francisco Parra:} Conceptualization, Methodology, Software, Validation,
Writing -- original draft.

\section*{Code availability section}
\noindent
Name of the code/library: \texttt{datacube-rs}\\
Contact: Francisco Parra, francisco.parra.o@usach.cl\\
Hardware requirements: any x86-64/ARM64 machine; benchmarks on a 16-core laptop\\
Program language: Rust (with Python bindings via PyO3 and a WebAssembly target)\\
Software required: Rust toolchain ($\geq$1.94, edition 2024)\\
Program size: TODO\\
Source code: \url{https://github.com/franciscoparrao/datacube-rs}\\
License: MIT OR Apache-2.0\\
Archived version: TODO (Zenodo DOI)

\section*{Data availability}
The Sentinel-2 case study uses freely available data from the Microsoft
Planetary Computer STAC catalogue. The synthetic series and scripts used for the
numerical-parity validation are included in the source repository.

\section*{Declaration of competing interest}
The author declares no competing interests.

\section*{Acknowledgements}
% TODO

\bibliography{refs}

\end{document}
